Amplitude metrics for cellular \\ circadian bioluminescence reporters

### 2014-09-02 response_letter.tex

Points 5 and 6 are linked concerns regarding the simulation and approximation of stochastic trajectories, so we have addressed them together.
The noise generated using the Gillespie algorithm for non-elementary propensity functions can indeed be different than the noise that would occur for the full system when expanded into elementary reactions.
That being said, the level of noise present in Fig.~5 was chosen by varying the volume parameter of the stochastic simulation so that desynchronization rates matched experimentally observed data.
We therefore do not believe that incorrect noise estimates from the Gillespie algorithm affects our approximations of the stochastic population.
{\itshape\bfseries Regardless, we tested the method with two additional models.}
First, the Oregonator, is a simple mass-action oscillatory reaction system considered in Gillespie's 1977 paper.
We used a perturbation to the conversion rate of two of the intermediate species at 6 different phases to test the accuracy of the perturbed approximation, $x_hat$.
Second, we tested the method with a stochastic simulation of a more detailed circadian model.
For consistency, we used the same model and perturbation as was used in Fig.~5 (for the first light pulse).
The results of these simulations are included as Fig.~S3, which (similar to Fig.~S2) demonstrates good agreement between the stochastic models and continuous approximations.

\begin{manuscript}{Supplemental Info, Page 6}
\bfseries
Figure S3: Validation of continuous methods with alternate models.
(\bfseries A) The Oregonator model (Model 3) was tested as an example of a mass-action limit cycle oscillator.
A population of 2000 oscillators was perturbed at size different mean phases, $\mu$, by increasing the $\mathit{c2}$ parameter by $50\%$ for $d=\pi$ (highlighted region).
Only the average $\mathit{Y2}$ variable is plotted.
({\bfseries B}) The model described in (2) (Model 2) was similarly tested at a variety of mean phases by reducing the $\mathit{vdp}$ parameter by $28.5\%$ and plotting the resulting $\mathit{c2}$ state variable (as in Fig.~5).
Parameters for the continuous approximations were found by estimating the phase, initial standard deviation, period, and phase diffusivity of the control population.
Good agreement is seen between the continuous and stochastic simulations, indicating the proposed method is suitable for predicting the population-level responses for a variety of model types.
\end{manuscript}

@@

\begin{reviewer}
Minor concerns:

1) In most case, ``hat'' mark is used to represent various perturbed variables.
However, ``hat'' mark is also used to represent the rescaled time variable.
This can lead to confusion in understanding the equations.
Please, use different notation for scaled time.
\end{reviewer}

This is a good suggestion, and {\itshape\bfseries we have changed the notation for scaled time to $\tilde{t}$ throughout the manuscript.}
Additionally, we noticed that our notation for pulse duration conflicted with our definition of phase diffusivity, so pulse duration has been changed to $\tilde{d}$.

@@

\begin{reviewer}
2) Please provide a little more detail for the derivation of (14).
\end{reviewer}

The derivation for Eq.~14 (now 11) is analogous to Eq.~28 (now 32), a lengthy process that was detailed in the appendix of Taylor et al, 2008 (PMID 19593456).
For space constraints, we have added only brief remarks on their derivation:

\begin{manuscript}{Page 5}
  Methods for efficiently calculating these quantities using ODE sensitivity analysis have been developed (24), with the important result that {\bfseries the parameter- and state-impulse PRCs can be related by the Jacobian matrix:}
\begin{equation}
  \frac{d}{d\tilde{t}}\frac{d\theta}{dp} = \frac{d\theta}{dx}\frac{d\tilde{f}}{dp} 
  \tag{11}
\end{equation}
{\bfseries This result follows from the fact} that in the limit {\bfseries of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.
This result follows from the fact that in the limit of an infinitely short and small parameter pulse, $\Delta x(0) \to \nicefrac{d\tilde{f}}{dp}\,\tilde{d}\, \Delta p$.}
\end{manuscript}

@@

{\bfseries\itshape This plot has been corrected.}

\begin{reviewer}
4) The PRC and ARC in Fig 4A are in response to which perturbation?
Please provide details on how Fig 4A is simulated.
\end{reviewer}

Additional details on the simulation of Fig. 4 have been added to both the main text and the figure legend:

\begin{manuscript}{Page 14}
{\bfseries We next demonstrate the usefulness of single-cell and population level ARCs in predicting the mean response of a population.
  Using a detailed single-cell model (Model 2), we calculate the response of the system, specifically the average CRY2 protein expression level, to a temporary increase in {\itshape Per} transcription rate.}
In Fig.~4A, we plot {\bfseries the resulting} single-cell and population-level response curves.
\end{manuscript}

\begin{manuscript}{Page 15}
  Figure 4: Response curves {\bfseries describe different types of amplitude change
A population of cells is simulated using a detailed model of circadian rhythms (Model 2).
Each cell in the population is subjected to a $20\%$ increase in the $\mathit{vtp}$ parameter ({\itshape Per} transcription rate) for $d=\nicefrac{\pi}{2}$, inducing both a phase and amplitude change.}
\end{manuscript}

@@

\begin{reviewer}
5) The definition of Eq (48) is not clear.
Please provide a more detailed explanation.
\end{reviewer}

Eq~ 48 (now 44) tracks the distance between the perturbed trajectory and the limit cycle for each phase.
The weighted-average of this distance is then used to find the population-level expression following perturbation.
To make this point more clear, we have added additional explanatory text to the manuscript:

\begin{manuscript}{Pages 12-13}
The second contribution to the perturbed population trajectory comes from deviations from limit cycle oscillations in each cell.
{\bfseries We calculate the population-level effect of these deviations by averaging over the deviations that occur at each phase.}
We define the deviation trajectory $\delta x(\theta, \tilde{t})$ for each phase as the distance between the perturbed trajectory and the phase-adjusted reference:
\begin{align}
  \delta x(\theta_0, \tilde{t}) &\coloneqq x(\tilde{t}) - x^\gamma(\tilde{t} + \theta_0 + \Delta \theta)\tag{48} \\
  \therefore\; \lim_{\tilde{t} \to \infty} \delta x(\theta, \tilde{t}) &= 0 \tag{49}
\end{align}
{\bfseries Since the perturbed trajectory ultimately converges with the phase-adjusted reference, deviations will converge to zero.
Since the phase change, $\Delta\theta$, associated with a perturbation at each phase is likely not known prior to calculating the perturbed trajectory, it is difficult to tabulate deviation trajectories associated with each final phase, $\theta_0 + \Delta\theta$.
It is therefore more straightforward to find the average effect of single-cell perturbations at the population level by weighting the deviations by the phase density function prior to perturbation.}
\end{manuscript}

@@

\begin{reviewer}
6) In a couple of places, the manuscript discusses pharmacological manipulation of circadian clocks.
Could you discuss this in more detail since the manuscript analyzes data coming from only light-induced perturbations (Fig.~5)? 
For instance, the manipulation of circadian rhythms with the CK1d/e inhibitor (PF-670462) has been studied widely.
Can you discuss how your method can be applied to the data in these studies?
\end{reviewer}

Indeed, while we would have preferred to analyze data coming from pharmacological perturbations, this type of data is difficult to obtain.
We have added an additional paragraph to the conclusion to address this limitation, and point out potential future applications of the method:

\begin{manuscript}{Page 18}
\bfseries
While our examples have focused on light-mediated perturbations, these approaches could also be applied to pharmacological perturbations.
Some difficulty arises in obtaining such data in cultured cells, however, as entraining perturbations would require that pharmacological agents be introduced for only a finite duration.
As medium or temperature changes are often enough to resynchronize cultured cells, a transient application is difficult to achieve experimentally.
The search for clock-enhancing molecules has therefore tended to focus on constant drug concentrations: for instance, dose-dependent period or amplitude change following inhibition of CK1$\delta$ or similar targets (48).
For {\itshape in vivo} systems, however, pharmacokinetics dictates a finite duration of action for both naturally secreted hormones and pharmacological therapies.
More effective treatments might therefore be designed by explicitly accounting for such a transient response, perturbing peripheral clocks at the right phase to induce resynchronization and an increase in single-cell level amplitudes.
\end{manuscript}

@@

We thank the reviewer for their positive impression of the manuscript, as well as for their helpful suggestions of related studies to discuss. We have incorporated many of the suggested works, which we feel helps to place the manuscript within the larger field of circadian modeling.

\begin{reviewer}
Detailed comments:

Felix Naef studied stochastic models to discuss fibroblast rhythms.
He found as well purely exponential decay in case of damped oscillators and deviations for limit cycles.
The current study should be compared to these models.
\end{reviewer}

In the relevant studies from Felix Naef, the authors use a phase model in which Gaussian white noise causes the frequencies to drift in time.
The decay rate is estimated from stochastic simulations of detailed limit cycle models, and used to capture unperturbed bioluminescence data.
We have incorporated two relevant references:

\begin{manuscript}{Page 21}
{\bfseries
32. Rougemont, J., and F. Naef, 2006. Collective synchronization in populations
of globally coupled phase oscillators with drifting frequencies. {\itshape Phys.\ Rev.\ E} 73:011104.\\}
\dots\\
{\bfseries
  36. Rougemont, J., and F. Naef, 2007. Dynamical signatures of cellular fluctuations and oscillator stability in peripheral circadian clocks. {\itshape Mol.\ Syst.\ Biol.\ } 3:93. }
\end{manuscript}

and added the references to the following sections of the manuscript:

\begin{manuscript}{Page 5}
\bfseries
Large populations of oscillators are typically described using phase-only models (32), in which the state of each oscillator is represented only by its phase, $\theta$.
\end{manuscript}
\begin{manuscript}{Page 6}
The phase diffusivity parameter $d$ (in units of inverse radians) describes the speed with which the population desynchronizes and can be fit to experimental data {\bfseries (36)}.
\end{manuscript}
\begin{manuscript}{Page 11}
\bfseries
Previous work has used limit cycle models to estimate population-level parameters, such as desynchronization rate, for phase-only models (36).
\end{manuscript}
\begin{manuscript}{Page 12}
  Phase-diffusion {\bfseries models can explain} why the gradual damping from experimental population-level data closely resembles an exponentially damped sinusoid, {\bfseries a result that has been shown experimentally (9) and computationally (36).}
\end{manuscript}

@@

In Fig.~2D, we plot the time-dependent state variables of the two-state model (Model 1).
These variables had not been explicitly plotted in this representation in another figure, so we wished to distinguish them from state-space representations or response curve variables.

\begin{reviewer}
David Rand published a generic theory of response functions which should include ARC?
A comparison is suggested. 
\end{reviewer}

Indeed, David Rand's generic theory of response functions would likely include our amplitude response curve as an example of a differentially changed quantity following a perturbation of finite duration.
Our contribution lies in the definition and derivation of the relevant formulas for amplitude change.
We have included the relevant citation from David Rand, along with a discussion in the manuscript text:

\begin{manuscript}{Page 3}
Mathematical models have long been used to understand the results of circadian experiments (15, 16), {\bfseries aided by} definitions and computational techniques {\bfseries designed to match modeling predictions to experimental data.
One such definition is the response function, a general technique that maps a change in an output variable to a temporary change in parameters (17).}
For instance, the phase response curve (PRC) has been used to characterize the entrainment behavior of both experimental and mathematical systems (18-20), and in analyzing the synchrony of populations of oscillators (21).
\end{manuscript}

@@

\begin{reviewer}
Joke Meijer (PLoS One) discussed the comparison of single cell amplitudes and ensembles of neuron regarding resetting dynamics as well. 
\end{reviewer}

Thank you for drawing our attention to this work, as it lends some experimental support to the bimodal distributions we observe following a light pulse.
We have incorporated the citation into the relevant section of the manuscript:

\begin{manuscript}{Page 14}
Changes to the synchronization of the population from each perturbation are readily apparent, {\bfseries with both perturbations inducing a bimodal distribution in the phase density.
While experimental data on individual cell phases was unavailable for this data set, bimodal distributions in SCN neuron firing following a phase shift have previously been seen experimentally (47).
As higher-frequency features, these bimodal distributions} dissipate as the phases diffuse.
\end{manuscript}

@@

\begin{reviewer}
The critical role of amplitudes for the response to perturbations has been discussed several times in the literature and a discussion of these previous studies can support the relevance of the current manuscript.
Examples: Pittendrigh JBR 1991, RK Barrett J.~Neuroscience 1995, MH Vitaterna PNAS 2006, S Brown PNAS 2008, A Granada PLoS One 2013, A.~Erzberger Proc Royal Soc 2013.

\end{reviewer}

The importance of amplitude in entrainment is an important consideration and has certainly been addressed by many studies in the past.
We have incorporated several of these references into the motivation for the manuscript, as designing therapies to boost circadian amplitudes will surely have an effect on the entrainment of the oscillators as well:

\begin{manuscript}{Page 2}
  {\bfseries The amplitude of circadian transcription is a relevant factor, and has been shown to play a critical role in phase resetting and entrainment (3, 4).}
Recent studies have further highlighted the importance of high peripheral clock amplitudes in maintaining metabolic health.
\end{manuscript}

@@

\begin{reviewer}
Ute Abraham et al. (Molecular Systems Biology) compare also coupled cells (SCN) with less coupled cells (lung) regarding their entrainment properties and discuss amplitude and relaxation effects.
\end{reviewer}

We have incorporated this reference into two places in the manuscript. First, in the introduction as reference (4) noted above, and also in the conclusion where we discuss the relative entrainment speed of coupled vs. uncoupled cells:

\begin{manuscript}{Page 18}
It is therefore interesting that liver cells have not developed a stronger mechanism of intercellular coupling, such as in the SCN, to maintain robust amplitudes in the absence
of external cues.
Perhaps the ability to quickly re-entrain to a phase shift, which is often slowed by coupling {\bfseries (4)}, has historically been more advantageous than protection against highly variable food intake.
\end{manuscript}
